---
title: "Comparative Tree Status"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: sandstone
    code_folding: hide
runtime: shiny
---

# Comparison of Various Tree Attributes

In this section

## Tree Attributes Relation to Eachother and Tree Attributes Distribution by Location

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(shiny)
library(rlang)
library(viridis)

theme_set(theme_minimal() + theme(legend.position = "right"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```


```{r load_data_define_function, message = FALSE}

# Load and wrangle data

trees_2015 <- read_csv("large_tree_data/2015_tree_raw.csv", na = c("", "NA", "Unknown")) |> 
  janitor::clean_names() %>% 
  mutate(spc_common = str_to_title(spc_common)) %>% 
  mutate(health = fct_relevel(health, c("Good", "Fair", "Poor")))

tree_status_df <- trees_2015 %>%
  select(curb_loc, guards, sidewalk, health, status,
         borough,
         root_stone, root_grate, root_other,
         trunk_wire, trnk_light, trnk_other,
         brch_light, brch_shoe, brch_other) %>% 
  mutate(health = fct_relevel(health, c("Good", "Fair", "Poor")),
         borough = as.factor(borough),
         All = as.factor("All"),
         curb_loc = case_match(curb_loc,
                               "OffsetFromCurb" ~ "Offset From Curb",
                               "OnCurb" ~ "On Curb"),
         curb_loc = as.factor(curb_loc),
         guards = as.factor(guards),
         sidewalk = case_match(sidewalk,
                               "NoDamage" ~ "No Damage",
                               "Damage" ~ "Damage"),
         sidewalk = as.factor(sidewalk),
         status = as.factor(status)
         ) %>%
  rename(`Curb Location` = curb_loc,
         `All of NYC` = All,
         Borough = borough,
         Guards = guards,
         Sidewalk = sidewalk,
         `Alive, Dead, or Stump` = status,
         Health = health,
         `Roots with Stones` = root_stone, 
         `Roots in Grate` = root_grate, 
         `Root Problem (Other)` = root_other,
         `Trunk has Wires` = trunk_wire, 
         `Trunk has Lights` = trnk_light, 
         `Trunk has Other Problem` = trnk_other,
         `Branches have Lights` = brch_light, 
         `Branches with Shoes` = brch_shoe, 
         `Branches Problem (Other)` = brch_other) %>% 
  select(Borough, `All of NYC`, everything())

exposure_options <- colnames(tree_status_df)[3:length(tree_status_df)]
outcome_options <- colnames(tree_status_df)[3:length(tree_status_df)]


# Define function

effective_tree_protection <- function(df, exposure_str, outcome_str){

  # Convert character strings to symbols
  exposure_sym <- rlang::sym(exposure_str)
  outcome_sym <- rlang::sym(outcome_str)

  # Make contingency table for the chi-square test
  contingency_table <- df %>%
    filter(!is.na(!!exposure_sym), !is.na(!!outcome_sym)) %>%
    group_by(!!outcome_sym, !!exposure_sym) %>%
    count(!!outcome_sym, !!exposure_sym) %>%
    pivot_wider(names_from = !!exposure_sym, values_from = n)

  # Perform the chi-square test
  chisq_test_result <- chisq.test(contingency_table[-1])
  chi_square <- chisq_test_result$statistic
  p_value <- chisq_test_result$p.value

  # Visualize the distribution differences 
  plot_title <- str_c("Distribution of ", exposure_str, 
                      " by ", outcome_str)

df %>%
    filter(!is.na(!!exposure_sym), !is.na(!!outcome_sym)) %>%
    group_by(!!exposure_sym, !!outcome_sym) %>%
    summarize(n = n(), .groups = 'drop') %>%
    group_by(!!exposure_sym) %>%
    mutate(Percent = n / sum(n) * 100) %>%
    ggplot(aes(x = !!exposure_sym, y = Percent, fill = !!outcome_sym)) +
    geom_bar(stat = 'identity', position = 'dodge') +
    geom_text(aes(label = sprintf("%.1f%%", Percent)),
              position = position_dodge(width = 0.9),
              size = 3.5, 
              vjust = -0.3) +
    labs(title = plot_title,
         subtitle = sprintf("p value = %.1e (Chi-square = %.2f)", p_value, chi_square)) 
}



```


### Tree Status in Different Scenarios

```{r input_}

# Input: Select variable for 'exposure'
selectInput(inputId = "exposure", 
         #   label = h3("Select Category (Exposure):"), 
            label = "Select Category (Exposure):", 
            choices = exposure_options,
            selected = "Curb Location"
            )

# Input: Select variable for 'outcome'
selectInput(inputId = "outcome", 
          #  label = h3("Select Tree Status (Outcome):"), 
            label = "Select Tree Status (Outcome):",
            choices = outcome_options,
            selected = "Health"
            )

```
  

```{r display_plot}

renderPlot({
    effective_tree_protection(df = tree_status_df, 
                              exposure_str = input$exposure, 
                              outcome_str = input$outcome)
  })


```

### Compare Tree Attributes in each Borough

```{r by_borough}

# Input: Select variable for 'outcome'
selectInput(inputId = "outcome_borough", 
            label = "Select Tree Status (Outcome):",
            choices = outcome_options,
            selected = "Health"
            )

renderPlot({
    effective_tree_protection(df = tree_status_df, 
                              exposure_str = "Borough", 
                              outcome_str = input$outcome_borough)
  })

```


### Look at Tree Attributes in NYC Overall

```{r overall}

# Input: Select variable for 'outcome'
selectInput(inputId = "outcome_all", 
            label = "Select Tree Status (Outcome):",
            choices = outcome_options,
            selected = "Health"
            )

renderPlot({
    effective_tree_protection(df = tree_status_df, 
                              exposure_str = "All of NYC", 
                              outcome_str = input$outcome_all)
  })

```


Tree Attributes in Different Scenarios

Several associations were explored between variables such as health, curb location, guards, sidewalk quality, and root, trunk, and branch quality (for example, if there are lights on the branches). There is an association between the location of the trees on the curb and whether branches have lights (trees on the curb are more likely to have lights on their branches). Additionally, there is an assoiation between sidewalk quality and the existence of root problems (damage on the sidewalk is associated with more root problems). 

```{r}

effective_tree_protection(df = tree_status_df, 
                              exposure_str = "Curb Location", 
                              outcome_str = "Branches have Lights")

effective_tree_protection(df = tree_status_df, 
                              exposure_str = "Sidewalk", 
                              outcome_str = "Root Problem (Other)")

```


Tree Attribute Distribution Overall

This figure shows the distribution of different attributes of trees and their status across all of NYC. For example, we see that 81.1% of trees are in “good” health, 14.8% of trees are in “fair” health, and 4.1% of trees are in “poor” health. Notably, 28.7% of sidewalks are damaged (71.2% not damaged). The majority of trees (87.8%) do not have guards and 20.5% of tree roots have stones. 

```{r}

# Apply function to relevant variables 

effective_tree_protection(df = tree_status_df, 
                              exposure_str = "All of NYC", 
                              outcome_str = "Health")

effective_tree_protection(df = tree_status_df, 
                              exposure_str = "All of NYC", 
                              outcome_str = "Sidewalk")

effective_tree_protection(df = tree_status_df, 
                              exposure_str = "All of NYC", 
                              outcome_str = "Roots with Stones")
```


Tree Health in Different Boroughs

Compared to the other four boroughs, Manhattan had the lowest percentage of trees in “good” health, and the highest percentage of trees in both “fair” and “poor” health. Overall in NYC, 81.1% of trees were in “good” health, 14.8% of trees were in “fair” health, and 4.1% were in “poor” health. 

```{r}

    effective_tree_protection(df = tree_status_df, 
                              exposure_str = "Borough", 
                              outcome_str = "Health")
```




