---
title: "Comparative Tree Status"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: sandstone
    code_folding: hide
---

# Comparison of Various Tree Attributes

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(shiny)
library(rlang)
library(viridis)
library(gridExtra)

theme_set(theme_minimal() + theme(legend.position = "right"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

```{r load_data_define_function, message = FALSE}

# Load and wrangle data

trees_2015 <- read_csv("large_tree_data/2015_tree_raw.csv", na = c("", "NA", "Unknown")) |> 
  janitor::clean_names() %>% 
  mutate(spc_common = str_to_title(spc_common)) %>% 
  mutate(health = fct_relevel(health, c("Good", "Fair", "Poor")))

tree_status_df <- trees_2015 %>%
  select(curb_loc, guards, sidewalk, health, status,
         borough,
         root_stone, root_grate, root_other,
         trunk_wire, trnk_light, trnk_other,
         brch_light, brch_shoe, brch_other) %>% 
  mutate(health = fct_relevel(health, c("Good", "Fair", "Poor")),
         borough = as.factor(borough),
         All = as.factor("All"),
         curb_loc = case_match(curb_loc,
                               "OffsetFromCurb" ~ "Offset From Curb",
                               "OnCurb" ~ "On Curb"),
         curb_loc = as.factor(curb_loc),
         guards = as.factor(guards),
         sidewalk = case_match(sidewalk,
                               "NoDamage" ~ "No Damage",
                               "Damage" ~ "Damage"),
         sidewalk = as.factor(sidewalk),
         status = as.factor(status)
         ) %>%
  rename(`Curb Location` = curb_loc,
         `All of NYC` = All,
         Borough = borough,
         Guards = guards,
         `Sidewalk Condition` = sidewalk,
         `Alive, Dead, or Stump` = status,
         Health = health,
         `Roots with Stones` = root_stone, 
         `Roots in Grate` = root_grate, 
         `Root Problem (Other)` = root_other,
         `Trunk has Wires` = trunk_wire, 
         `Trunk has Lights` = trnk_light, 
         `Trunk Problem (Other)` = trnk_other,
         `Branches have Lights` = brch_light, 
         `Branches with Shoes` = brch_shoe, 
         `Branches Problem (Other)` = brch_other) %>% 
  select(Borough, `All of NYC`, everything())

exposure_options <- colnames(tree_status_df)[3:length(tree_status_df)]
outcome_options <- colnames(tree_status_df)[3:length(tree_status_df)]

# Define function for plot

tree_attribute_plot <- function(df, exposure_str, outcome_str){

  # Convert character strings to symbols
  exposure_sym <- rlang::sym(exposure_str)
  outcome_sym <- rlang::sym(outcome_str)

  # Make contingency table for the chi-square test
  contingency_table <- df %>%
    filter(!is.na(!!exposure_sym), !is.na(!!outcome_sym)) %>%
    group_by(!!outcome_sym, !!exposure_sym) %>%
    count(!!outcome_sym, !!exposure_sym) %>%
    pivot_wider(names_from = !!exposure_sym, values_from = n)

  # Perform the chi-square test
  chisq_test_result <- chisq.test(contingency_table[-1])
  chi_square <- chisq_test_result$statistic
  p_value <- chisq_test_result$p.value

  # Visualize the distribution differences 

df %>%
    filter(!is.na(!!exposure_sym), !is.na(!!outcome_sym)) %>%
    group_by(!!exposure_sym, !!outcome_sym) %>%
    summarize(n = n(), .groups = 'drop') %>%
    group_by(!!exposure_sym) %>%
    mutate(Percent = n / sum(n) * 100) %>%
    ggplot(aes(x = !!exposure_sym, y = Percent, fill = !!outcome_sym)) +
    geom_bar(stat = 'identity', position = 'dodge') +
    geom_text(aes(label = sprintf("%.1f%%", Percent)),
              position = position_dodge(width = 0.9),
              size = 3.5, 
              vjust = -0.3) +
    labs(title = str_c(outcome_str, " by ", exposure_str),
         subtitle = sprintf("p value = %.1e (Chi-square = %.2f)", p_value, chi_square)) +
    theme(legend.position="bottom")
}

# Define layout for tabs

exposure_tab <- function(exposure){
  
grid.arrange(
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones"),
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate"),
  ncol = 2)
grid.arrange(
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)"),
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)"),
  ncol = 2)
grid.arrange(
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires"),
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights"),
  ncol = 2)
grid.arrange(
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones"),
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate"),
  ncol = 2)
grid.arrange(
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights"),
  tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)"),
  ncol = 2)

tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Alive, Dead, or Stump")
  
}


```


# Tree Attributes Relation to Eachother and Tree Attributes Distribution by Location 

# Tree Status in Different Scenarios {.tabset}

## Curb Location {.tabset}

```{r}
exposure <- "Curb Location"
```

### Tree Health

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Health")
```


### Roots
```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)")
```

### Trunk

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)")
```

### Branches

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)")
```

### Vital Status

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Alive, Dead, or Stump")
```


## Sidewalk Condition {.tabset}

```{r}
exposure <- "Sidewalk Condition"
```

### Tree Health {.tabset}

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Health")
```


### Roots
```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)")
```

### Trunk

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)")
```

### Branches

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)")
```

### Vital Status

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Alive, Dead, or Stump")
```


## Tree Guards {.tabset}

```{r}
exposure <- "Guards"
```

### Tree Health {.tabset}

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Health")
```


### Roots
```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)")
```

### Trunk

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)")
```

### Branches

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)")
```

### Vital Status

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Alive, Dead, or Stump")
```

## Tree Health {.tabset}

```{r}
exposure <- "Health"
```

### Roots {.tabset}
```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)")
```

### Trunk

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)")
```

### Branches

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)")
```

# Tree Attributes Distribution by Borough and Overall in NYC {.tabset}

## Distribution by Borough {.tabset}

```{r}
exposure <- "Borough"
```

### Tree Health {.tabset}

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Health")
```

### Tree Location 

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Sidewalk Condition")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Curb Location")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Guards")
```

### Roots
```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)")
```

### Trunk

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)")
```

### Branches

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)")
```

### Vital Status

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Alive, Dead, or Stump")
```

## Overall NYC

```{r}
exposure <- "All of NYC"
```

### Tree Health {.tabset}

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Health")
```

### Tree Location 

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Sidewalk Condition")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Curb Location")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Guards")
```

### Roots
```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots with Stones")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Roots in Grate")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Root Problem (Other)")
```

### Trunk

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Wires")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk has Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Trunk Problem (Other)")
```

### Branches

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches have Lights")
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Branches Problem (Other)")
```

### Vital Status

```{r}
tree_attribute_plot(df = tree_status_df, exposure_str = exposure, outcome_str = "Alive, Dead, or Stump")
```
