---
title: "Zander's Brainstorming Page"
---

### Tree Distribution, Ecological Variables, and Health Relationships

```{r Page Settings, message = FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)
library(plotly)
library(sf)
library(tmap)
library(tmaptools)
library(readxl)

theme_set(theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r Exploring DBH Values}

#trees_2015_cleaned = read_csv("large_tree_data/tree_2015_cleaned.csv")

trees_2015_cleaned = read_csv("large_tree_data/2015_tree_raw.csv", na = c("", "NA", "Unknown")) |> 
  janitor::clean_names()

trees_dbh = trees_2015_cleaned |> 
  drop_na(tree_dbh) |> 
  group_by(borough, nta_name) |> 
  summarize(
    n_trees = n(),
    avg_dbh = mean(tree_dbh)) |> 
  arrange(desc(avg_dbh))

large_tree_table = trees_dbh |> 
  ungroup() |> 
  filter(n_trees > 100) |> 
  filter(min_rank(desc(avg_dbh)) < 11) |> 
  knitr::kable(digits = 2, caption= "Top 10 Neighborhoods wth the Highest Average DBH (inches) Across Local Street Trees")

large_tree_table

```

```{r Borough DBH Boxplot}
borough_dbh = trees_2015_cleaned |> 
  drop_na(tree_dbh) |> 
  group_by(borough) |> 
  mutate(borough_dbh = mean(tree_dbh))

borough_dbh |> 
  filter(tree_dbh < 101) |> 
ggplot(aes(x = borough, y = tree_dbh, fill = borough)) + geom_boxplot() + scale_y_continuous(breaks = scales::pretty_breaks(10)) +
  labs(
    title = "Average Tree DBH Across NYC Boroughs",
    y = "Measured DBH (Inches)",
    x = "Borough"
  )
  
```

```{r Plotly Version DBH Boxplot}
borough_dbh |> 
  filter(tree_dbh < 101) |> 
  plot_ly(y = ~tree_dbh, color = ~borough, type = "box", colors = "viridis")
```


```{r Outlier Trees}
#70 Trees have a dbh higher than 100 inches, removed from visualization above to see boxplots better
outlier_trees = borough_dbh |> 
  filter(tree_dbh > 101) |> 
  count(tree_dbh)

outlier_trees |> 
  knitr::kable()
```

```{r Initial Histogram}
borough_dbh |> 
  filter(tree_dbh < 51) |> 
   ggplot(aes(x = tree_dbh)) + geom_histogram() + scale_y_continuous(breaks = scales::pretty_breaks(10)) + scale_x_continuous(breaks = scales::pretty_breaks(10)) + 
   labs(
    title = "Overall Distribution of Tree Diameters at Breast Height, New York City",
    x = "Measured DBH (Inches)",
    y = "Number of Trees"
  )
```

```{r Density Plot Faceted By Borough}

borough_dbh |> 
  filter(tree_dbh < 50) |> 
  ggplot(aes(x = tree_dbh, fill = borough)) + 
  geom_density() + facet_grid(borough ~ .) +
  scale_x_continuous(breaks = scales::pretty_breaks(12)) +
  labs(
    title = "Density Plot of All Measured Tree Diameters Below 50 inches",
    x = "Measured DBH (Inches)",
    y = "Density"
  )


```

```{r NTA Average DBH }
neighborhood_dbh = trees_2015_cleaned |> 
  drop_na(tree_dbh) |> 
  group_by(nta_name) |> 
  mutate(nta_dbh = mean(tree_dbh)) |> 
  ungroup() |> 
  mutate(nta_name = fct_reorder(nta, nta_dbh))

neighborhood_dbh |> 
ggplot(aes(x = nta_name, y = nta_dbh, color = borough)) + geom_point() + facet_grid(borough ~ .) + theme(axis.text.x = element_text(angle = 60)) +
  labs(
    title = "Average DBH By NYC Neighborhood Tabulation Area",
    x = "NTA Neighborhood",
    y = "Mean DBH. (Inches)"
  )

```


# Spatially Mapping Neighborhood DBH Values Across NYC

```{r Importing NTA Shapefile for Data Join}
#Reading in Shapefile
nyc_nta = st_read("small_data/geo_export_e924b274-8b6e-427d-87c0-92bfde8ce30a.shp")

glimpse(nyc_nta)
head(nyc_nta)

tm_shape(nyc_nta) +
  tm_polygons()

```

```{r Creating DBH Join layer as new small dataframe, Joining}

nta_dbh_summarized = trees_2015_cleaned |> 
  drop_na(tree_dbh) |> 
  group_by(nta) |> 
  summarize(
    "Number of Trees" = n(),
    avg_dbh = mean(tree_dbh))


nta_dbh_spatial = merge(nyc_nta, nta_dbh_summarized, 
                        by.x = "ntacode",
                        by.y = "nta")
```

```{r Visualizing DBH Data Test 1}

#Fast to render using the summarized df instead of full tree csv dataframe, but noticing some NTA's are missing from the Street Tree Data Census 188/195 are present

#Missing NTA Values appear to be park green spaces? How can I fix or color them a different shade. Is it fine to have these gaps?

tm_shape(nta_dbh_spatial) +
  tm_polygons(col = "avg_dbh",
              style = "quantile",
              n = 5,
              palette = "YlGn",
              border.col = "black",
              title = "Mean Tree DBH (Inches)") +
  tm_layout(main.title = "Average Tree DBH Across NYC Neighborhoods",
            title.size = 0.5) +
  tm_legend(legend.position = c("left","center"))
```

### Air Pollutants and Tree Distribution
The EPA and Leading Forest Agencies discuss that one of the ecosystem benefits of urban street trees is the ability for trees to uptake fine particulate matter (PM2.5) from vehicle exhaust and local combustion through stomata uptake during photosynthesis. https://enviroatlas.epa.gov/enviroatlas/DataFactSheets/pdf/ESC/PercentparticulatematterPM25removedannuallybytreecover.pdf

This content can be used in conjuction with Stephanie's Asthma and Tree Distribution data analysis

PM2.5 2011-2014 Projection Data From:
New York City Department of Health, Environment & Health Data Portal. Air quality data. Fine particles (PM 2.5). Accessed at https://a816-dohbesp.nyc.gov/IndicatorPublic/data-explorer/air-quality/ on 12/03/2023.
These are on the level of Community District

```{r Community District}
#Reading in Shapefile
nyc_comdist = st_read("small_data/geo_export_bd3c92bc-94de-48f3-9de3-e939473f307b.shp")

glimpse(nyc_comdist)
head(nyc_comdist)

tm_shape(nyc_comdist) +
  tm_polygons()
```

```{r Organizing PM2.5 and Tree Distribution 2015 Data for merging}

nyc_finepm2015 = read_csv("small_data/nyc_pm2.5_annual2015.csv") |> 
  janitor::clean_names() |> 
  filter(time == "Annual Average 2015" & geo_type_desc == "Community District") |> 
  select(geo_id, geography, avg_pm2.5 = mean_mcg_m3)
  

#In trees_2015_cleaned, community district is under "community board" id number
trees_cd_summarized = trees_2015_cleaned |> 
  drop_na(tree_dbh) |> 
  mutate(health = as.factor(health)) |> 
  group_by(community_board) |> 
  summarize(
    "Number of Trees" = n(),
    avg_dbh = mean(tree_dbh),
    health_status = mode(health))
```

```{r Merging Pm2.5 and Tree Data }

nyc_comdistrict_spatial = merge(nyc_comdist, trees_cd_summarized, nyc_finepm2015,
                        by.x = "boro_cd",
                        by.y = "community_board")

nyc_comdistrict_treesPM = merge(nyc_comdistrict_spatial, nyc_finepm2015,
                        by.x = "boro_cd",
                        by.y = "geo_id")
```

```{r PM2.5 Map and Number of Trees}
tm_shape(nyc_comdistrict_treesPM) +
  tm_polygons(col = "avg_pm2.5",
              style = "quantile",
              n = 5,
              palette = "YlOrBr",
              border.col = "black",
              title = "Average PM2.5 Concentration (µg/m3)") +
tm_shape(nyc_comdistrict_treesPM) +
  tm_dots(col = "black",
          size = "Number of Trees",
          style = "quantile",
          border.col = "black") +
  tm_layout(main.title = "Average 2015 PM2.5 Concentrations across NYC Community Districts", main.title.size = 0.85) +
  tm_legend(legend.position = c("left","top"))
```

```{r Pm2.5 but switching out Average Tree DBH Size }

#Doesnt work very well at this scale.
tm_shape(nyc_comdistrict_treesPM) +
  tm_polygons(col = "avg_pm2.5",
              style = "quantile",
              n = 5,
              palette = "YlOrBr",
              border.col = "black",
              title = "Average PM2.5 Concentration (µg/m3)") +
tm_shape(nyc_comdistrict_treesPM) +
  tm_dots(col = "black",
          size = "avg_dbh",
          style = "quantile",
          border.col = "black") +
  tm_layout(main.title = "Average 2015 PM2.5 Concentrations across NYC Community Districts", main.title.size = 0.8) +
  tm_legend(legend.position = c("left","top"))
```

## Heat Stress Available Data
How do trees impact the urban heat island effect through shading and heatstress events like hospitalizations?

```{r Heat Stress Data and Merger}

nyc_heatstress_2015 = read_csv("small_data/nyc_Heatstress Hospitalizations_2011-2015.csv", na = c(" ", "NA", "***")) |> 
  janitor::clean_names() |> 
  filter(geo_type_desc == "Community District") |> 
  mutate(
    average_annual_age_adjusted_rate_per_100_000 = 
      str_replace(average_annual_age_adjusted_rate_per_100_000, "[*]$", "")) |>
  select(geo_id, avg_heat_hospit = average_annual_age_adjusted_rate_per_100_000, heat_hospitalizations = total) |> 
  mutate(avg_heat_hospit = as.numeric(avg_heat_hospit))

nyc_comdistrict_heatstress = merge(nyc_comdistrict_treesPM, nyc_heatstress_2015,
                        by.x = "boro_cd",
                        by.y = "geo_id")

```

```{r Heat Stress Map}

tm_shape(nyc_comdistrict_heatstress) +
  tm_polygons(col = "avg_heat_hospit",
              style = "jenks",
              n = 5,
              palette = "OrRd",
              border.col = "black",
              title = "Avg. Adjusted Hospitalization Rate per 100k") +
tm_shape(nyc_comdistrict_treesPM) +
  tm_dots(col = "black",
          size = "Number of Trees",
          style = "quantile",
          border.col = "black") +
  tm_layout(main.title = "Average Heatstress Hospitalization Rate across NYC Community Districts", main.title.size = 0.8) +
  tm_legend(legend.position = c("left","top"))
```

```{r Asthma Hospitalization Merge}
asthma_hosp_adult_raw <-
  read_csv("small_data/NYC EH Data Portal - Asthma hospitalizations adults.csv") |> 
  janitor::clean_names()

asthma_hosp_adult_clean <- asthma_hosp_adult_raw |> 
  rename(nta_name = geography) |> 
  filter(geo_type == "NTA2010" & time == "2012-2014") |> 
  select(nta_name, geo_id, average_annual_age_adjusted_rate_per_10_000, average_annual_rate_per_10_000)

nta_asthma_spatial = merge(nta_dbh_spatial, asthma_hosp_adult_clean, 
                        by.x = "ntaname",
                        by.y = "nta_name")
```

```{r Asthma Hospitalization Map Health Page}
tm_shape(nta_asthma_spatial) +
  tm_polygons(col = "average_annual_rate_per_10_000",
              style = "quantile",
              n = 5,
              palette = "Purples",
              border.col = "black",
              title = "Asthma Hospitalization Rate per 10k") +
  tm_layout(main.title = "Avg. Annual Hospitalization Rate Across NYC Neighborhoods", main.title.size = 1) +
  tm_legend(legend.position = c("left","top"))
```

```{r Tree Count Map to Compare to Asthma }
tm_shape(nta_asthma_spatial) +
  tm_polygons(col = "Number of Trees",
              style = "quantile",
              n = 5,
              palette = "Greens",
              border.col = "black",
              title = "Number of Trees per NTA") +
  tm_layout(main.title = "Total Tree Count Across NYC Neighborhoods", main.title.size = 1) +
  tm_legend(legend.position = c("left","top"))
```

